Activity 6 - State-space models
========================================================

This activity will explore the state-space framework for modeling time-series and spatial data sets. Chapter 8 provides a more in-depth description of the state-space model, but in a nutshell it is based on separating the process model, which describes how the system evolves in time or space, from the observation error model. Furthermore, the state-space model gets its name because the model estimates that true value of the underlying **latent** state variables.

For this activity we will write all the code, process all the data, and visualize all the outputs in R, but the core of the Bayesian computation will be handled by JAGS (Just Another Gibbs Sampler, http://mcmc-jags.sourceforge.net). Therefore, before we get started you will want to download both the JAGS software and the rjags library, which allows R to call JAGS.

```{r}
library(rjags)
```

Next we'll want to grab the data we want to analyze. For this example we'll use the Google Flu Trends data for the state of Massachusetts, which we saw how to pull directly off the web in Activity 3.

```{r}
gflu = read.csv("http://www.google.org/flutrends/about/data/flu/us/data.txt",skip=11)
time = as.Date(gflu$Date)
y = gflu$Massachusetts
plot(time,y,type='l',ylab="Flu Index",lwd=2,log='y')
```

Next we'll want to define the JAGS code, which we'll do this by writing the code as a string in R. The code itself has three components, the data model, the process model, and the priors. The data model relates the observed data, y, at any time point to the latent variable, x. For this example we'll assume that the observation model just consists of Gaussian observation error. The process model relates the state of the system at one point in time to the state one time step ahead. In this case we'll start with the simplest possible process model, a random walk, which just consists of Gaussian process error centered around the current value of the system. Finally, for the priors we need to define priors for the initial condition, the process error, and the observation error.

```{r}
RandomWalk = "
model{
  
  #### Data Model
  for(i in 1:n){
    y[i] ~ dnorm(x[i],tau_obs)
  }
  
  #### Process Model
  for(i in 2:n){
    x[i]~dnorm(x[i-1],tau_add)
  }
  
  #### Priors
  x[1] ~ dnorm(x_ic,tau_ic)
  tau_obs ~ dgamma(a_obs,r_obs)
  tau_add ~ dgamma(a_add,r_add)
}
"
```

Next we need to define the data and priors as a list. For this analysis we'll work with the log of the Google flu index since the zero-bound on the index and the magnitudes of the changes appear much closer to a log-normal distribution than to a normal.
```{r}
data <- list(y=log(y),n=length(y),x_ic=log(1000),tau_ic=100,a_obs=1,r_obs=1,a_add=1,r_add=1)
```

Next we need to definite the initial state of the model's parameters for each chain in the MCMC. The overall initialization is stored as a list the same length as the number of chains, where each chain is passed a list of the initial values for each parameter. Unlike the definition of the priors, which had to be done independent of the data, the inidialization of the MCMC is allowed (and even encouraged) to use the data. However, each chain should be started from different initial conditions. We handle this below by basing the initial conditions for each chain off of a different random sample of the original data. 
```{r}
nchain = 3
init <- list()
for(i in 1:nchain){
  y.samp = sample(y,length(y),replace=TRUE)
  init[[i]] <- list(tau_add=1/var(diff(log(y.samp))),tau_obs=5/var(log(y.samp)))
}
```

Now that we've defined the model, the data, and the initialization, we need to send all this info to JAGS, which will return the JAGS model object.
```{r}
j.model   <- jags.model (file = textConnection(RandomWalk),
                             data = data,
                             inits = init,
                             n.chains = 3)
```

Next, given the defined JAGS model, we'll want to take a few samples from the MCMC chain and assess when the model has converged. To take samples from the MCMC object we'll need to tell JAGS what variables to track and how many samples to take.
```{r}
## burn-in
jags.out   <- coda.samples (model = j.model,
                            variable.names = c("tau_add","tau_obs"),
                                n.iter = 1000)
plot(jags.out)
```

Here we see that the model converges rapidly. Since rjags returns the samples as a CODA object, we can use any of the diagnositics in the R *coda* library to test for convergence, summarize the output, or visualize the chains.

Now that the model has converged we'll want to take a much larger sample from the MCMC and include the full vector of X's in the output
```{r}
jags.out   <- coda.samples (model = j.model,
                            variable.names = c("x","tau_add","tau_obs"),
                                n.iter = 10000)
```

Given the full joint posteror samples, we're next going to visualize the output by just looking at the 95% credible interval of the timeseries of X's and compare that to the observed Y's. To do so we'll convert the coda output into a matrix and then calculate the quantiles. Looking at colnames(out) will show you that the first two columns are `tau_add` and `tau_obs`, so we calculate the CI starting from the 3rd column. We also transform the samples back from the log domain to the linear domain.
```{r}
time.rng = c(1,length(time)) ## adjust to zoom in and out
ciEnvelope <- function(x,ylo,yhi,...){
  polygon(cbind(c(x, rev(x), x[1]), c(ylo, rev(yhi),
                                      ylo[1])), border = NA,...) 
}
out <- as.matrix(jags.out)
ci <- apply(exp(out[,3:ncol(out)]),2,quantile,c(0.025,0.5,0.975))

plot(time,ci[2,],type='n',ylim=range(y,na.rm=TRUE),ylab="FluIndex",log='y',xlim=time[time.rng])
## adjust x-axis label to be monthly if zoomed
if(diff(time.rng) < 100){ 
  axis.Date(1, at=seq(time[time.rng[1]],time[time.rng[2]],by='month'), format = "%Y-%m")
}
ciEnvelope(time,ci[1,],ci[3,],col="lightBlue")
points(time,y,pch="+",cex=0.5)
```

Next, lets look at the posterior distributions for `tau_add` and `tau_obs`, which we'll convert from precisions back into standard deviations. We'll also want to look at the joint distribution of the two parameters to check whether the two parameters strongly covary.
```{r}
layout(matrix(c(1,2,3,3),2,2,byrow=TRUE))
hist(1/sqrt(out[,1]),main=colnames(out)[1])
hist(1/sqrt(out[,2]),main=colnames(out)[2])
plot(out[,1],out[,2],pch=".",xlab=colnames(out)[1],ylab=colnames(out)[2])
cor(out[,1:2])
```

Assignment:
-----------

To look at how observation frequency affects data assimilation, convert 3 out of every 4 observations to NA (i.e. treat the data as approximately monthly) and refit the model. 

* Generate a time-series plot for the CI of x that includes the observations (as above). Use a different color and symbol to differentiate observations that were included in the model versus those that were converted to NA's.
```{r}
library(rjags)

gflu = read.csv("http://www.google.org/flutrends/about/data/flu/us/data.txt",skip=11)
time = as.Date(gflu$Date)
y = gflu$Massachusetts
plot(time,y,type='l',ylab="Flu Index",lwd=2,log='y')

even<-seq(2,620, by=2)
fourth<-seq(1,620, by=4)
need.replace<-c(even, fourth)

time2<- replace(gflu$Date,need.replace,NA)
time2_rem<-na.omit(time2)
time2_date=as.Date(time2_rem)

y.new<-replace(gflu$Massachusetts,need.replace,NA)
y.new<-as.numeric(na.omit(y.new))
plot(time2_date,y.new,type='l',ylab="Flu Index",lwd=2,log='y')


RandomWalk = "
model{

#### Data Model
for(i in 1:n){
y.new[i] ~ dnorm(x[i],tau_obs)
}

#### Process Model
for(i in 2:n){
x[i]~dnorm(x[i-1],tau_add)
}

#### Priors
x[1] ~ dnorm(x_ic,tau_ic)
tau_obs ~ dgamma(a_obs,r_obs)
tau_add ~ dgamma(a_add,r_add)
}
"

data <- list(y.new=log(y.new),n=length(y.new),x_ic=log(1000),tau_ic=100,a_obs=1,r_obs=1,a_add=1,r_add=1)

nchain = 3
init <- list()
for(i in 1:nchain){
  y.samp.new = sample(y.new,length(y.new),replace=TRUE)
  init[[i]] <- list(tau_add=1/var(diff(log(y.samp.new)),na.rm=TRUE),tau_obs=5/var(log(y.samp.new),na.rm=TRUE))
}

j.model   <- jags.model (file = textConnection(RandomWalk),
                         data = data,
                         inits = init,
                         n.chains = 3)

## burn-in
jags.out   <- coda.samples (model = j.model,
                            variable.names = c("tau_add","tau_obs"),
                            n.iter = 1000)
plot(jags.out)

jags.out   <- coda.samples (model = j.model,
                            variable.names = c("x","tau_add","tau_obs"),
                            n.iter = 10000)

time.rng = c(1,length(time2_date)) ## adjust to zoom in and out
ciEnvelope <- function(x,ylo,yhi,...){
  polygon(cbind(c(x, rev(x), x[1]), c(ylo, rev(yhi),
                                      ylo[1])), border = NA,...) 
}
out.new <- as.matrix(jags.out)
ci.new <- apply(exp(out.new[,3:ncol(out.new)]),2,quantile,c(0.025,0.5,0.975))

plot(time2_date,ci.new[2,],type='n',ylim=range(y.new,na.rm=TRUE),ylab="Flu Index",log='y',xlim=time2_date[time.rng])




## adjust x-axis label to be monthly if zoomed
if(diff(time.rng) < 100){ 
  axis.Date(1, at=seq(time2_date[time.rng[1]],time2_date[time.rng[2]],by='month'), format = "%Y-%m")
}

ciEnvelope(time2_date,ci.new[1,],ci.new[3,],col="lightBlue")

points(time2_date,y.new,pch="+",cex=0.8,col="red")

points(time,y,pch="-",cex=0.5)
```

* Compare the CI between the two runs.
```{r}
plot(time2_date,ci.new[2,],type='n',ylim=range(y.new,na.rm=TRUE),ylab="Flu Index",log='y',xlim=time2_date[time.rng],xlab="Time")
ciEnvelope(time2_date,ci.new[1,],ci.new[3,],col="lightBlue")

RandomWalk = "
model{

#### Data Model
for(i in 1:n){
y[i] ~ dnorm(x[i],tau_obs)
}

#### Process Model
for(i in 2:n){
x[i]~dnorm(x[i-1],tau_add)
}

#### Priors
x[1] ~ dnorm(x_ic,tau_ic)
tau_obs ~ dgamma(a_obs,r_obs)
tau_add ~ dgamma(a_add,r_add)
}
"
data <- list(y=log(y),n=length(y),x_ic=log(1000),tau_ic=100,a_obs=1,r_obs=1,a_add=1,r_add=1)
init_old <- list()
for(i in 1:nchain){
  y.samp_old = sample(y,length(y),replace=TRUE)
  init_old[[i]] <- list(tau_add=1/var(diff(log(y.samp_old))),tau_obs=5/var(log(y.samp_old)))
}
j.model_old   <- jags.model (file = textConnection(RandomWalk),
                         data = data,
                         inits = init_old,
                         n.chains = 3)
jags.out_old   <- coda.samples (model = j.model_old,
                            variable.names = c("x","tau_add","tau_obs"),
                            n.iter = 10000) 
out_old <- as.matrix(jags.out_old)
ci_old <- apply(exp(out_old[,3:ncol(out_old)]),2,quantile,c(0.025,0.5,0.975))

ciEnvelope(time,ci_old[1,],ci_old[3,],col="grey")

```
```
The CI of the orginal plot, which had more data points, was tighter, as makes sense with better data.

```
* Generate a predicted (median) vs observed plot for the data points that were removed
##plot obs(y) on x axis and predicted median(ci[2,]) on yaxis
```{r}
used<-seq(1,620, by=4)

y.unused<-replace(gflu$Massachusetts,used,NA)
y.unused<-as.numeric(na.omit(y.unused))
init_unused <- list()
for(i in 1:nchain){
  y.samp.unused = sample(y.unused,length(y.unused),replace=TRUE)
  init_unused[[i]] <- list(tau_add=1/var(diff(log(y.samp.unused)),na.rm=TRUE),tau_obs=5/var(log(y.samp.unused),na.rm=TRUE))
}
RandomWalk = "
model{

#### Data Model
for(i in 1:n){
y.unused[i] ~ dnorm(x[i],tau_obs)
}

#### Process Model
for(i in 2:n){
x[i]~dnorm(x[i-1],tau_add)
}

#### Priors
x[1] ~ dnorm(x_ic,tau_ic)
tau_obs ~ dgamma(a_obs,r_obs)
tau_add ~ dgamma(a_add,r_add)
}
"
data <- list(y.unused=log(y.unused),n=length(y.unused),x_ic=log(1000),tau_ic=100,a_obs=1,r_obs=1,a_add=1,r_add=1)
j.model_unused   <- jags.model (file = textConnection(RandomWalk),
                         data = data,
                         inits = init_unused,
                         n.chains = 3)
jags.out_unused   <- coda.samples (model = j.model_unused,
                            variable.names = c("x","tau_add","tau_obs"),
                            n.iter = 10000)
out.unused <- as.matrix(jags.out_unused)
ci.unused <- apply(exp(out.unused[,3:ncol(out.unused)]),2,quantile,c(0.025,0.5,0.975))

plot(y.unused,ci.unused[2,],xlab="observed values", ylab="predicted values")
abline(0,1, cex=0.5)

```
* Comment on the accuracy and precision of the estimates.
```
The accuracy and precision are both very high. This may have to do with the fact that there were so many data points to work with

```
Extra Credit:
-------------

Return to the original data and instead of removing 3/4 of the data remove the last 40 observations (convert to NA) and refit the model to make a forecast for this period

* Generate a time-series plot for the CI of x that includes the observations (as above but zoom the plot on the last ~80 observations). Use a different color and symbol to differentiate observations that were included in the model versus those that were converted to NA's.
* Comment on how well the random walk model performed (both accuracy and precision) and how it might be modified to improve both these criteria.